---
title: "analysis"
author: "Zakari L. Billo"
output: pdf_document
---

```{r include=false}
library(tidyverse)
library(glmnet)
library(pcalg)
library(foreach)
library(doParallel)
library(reshape2)
library(graph)
library(stats)
```

## Load clean datasets

```{r}
X_filtered <- read_csv("data/X_filtered.csv")
X_filtered <- as.data.frame(X_filtered)

X_asthma <- read_csv("data/X_asthma.csv")
X_asthma <- as.data.frame(X_asthma)

X_normal <- read_csv("data/X_normal.csv")
X_normal <- as.data.frame(X_normal)
```

Run Lasso for the "HLA-DQB1" neighborhood and Compare lamda options for optimal selection

```{r}
y <- as.numeric(X_filtered[, "HLA-DQB1"])
x <- as.matrix(X_filtered[, setdiff(colnames(X_filtered), "HLA-DQB1")])

# Test alphas from 0.1 to 1.0. 
alpha_seq <- seq(0.1, 1, by = 0.1)
cv_results <- lapply(alpha_seq, function(a) {
  cv.glmnet(x, y, alpha = a, nfolds = 10) 
})

# Alpha with minimum mean cross-validated error
min_errors <- sapply(cv_results, function(cv) min(cv$cvm))
best_alpha_idx <- which.min(min_errors)
best_alpha <- alpha_seq[best_alpha_idx]

print(paste("Optimal Alpha found:", best_alpha))

# Run Model with criteria
full_fit <- glmnet(x, y, alpha = best_alpha)
lambda_seq <- full_fit$lambda

n <- nrow(x)
p <- ncol(x)

# RSS Calculation
RSS <- (1 - full_fit$dev.ratio) * full_fit$nulldev
df <- full_fit$df

# Criteria
aic <- n * log(RSS / n) + 2 * df
bic <- n * log(RSS / n) + log(n) * df

# EBIC
gamma <- 0.5
ebic <- bic + 2 * gamma * lchoose(p, df)

# Optimal lambdas
opt_lambda_aic  <- lambda_seq[which.min(aic)]
opt_lambda_bic  <- lambda_seq[which.min(bic)]
opt_lambda_ebic <- lambda_seq[which.min(ebic)]
```

Visualization for Lasso approach

```{r}
criteria_df <- data.frame(
  LogLambda = log(lambda_seq),
  AIC = scale(aic), 
  BIC = scale(bic),
  EBIC = scale(ebic)
)

criteria_long <- melt(criteria_df, id.vars = "LogLambda")

p1 <- ggplot(criteria_long, aes(x = LogLambda, y = value, color = variable)) +
  geom_line(size = 1) +
  geom_vline(xintercept = log(opt_lambda_ebic), linetype="dashed", color="darkgreen") +
  annotate("text", x=log(opt_lambda_ebic), y=max(criteria_long$value), 

           label="EBIC", hjust=-0.1, color="darkgreen") +
  geom_vline(xintercept = log(opt_lambda_bic), linetype="dashed", color="orange") +
  annotate("text", x=log(opt_lambda_bic), y=max(criteria_long$value), 

           label="BIC", hjust=1.1, color="orange") +
  geom_vline(xintercept = log(opt_lambda_aic), linetype="dashed", color="purple") +
  annotate("text", x=log(opt_lambda_aic), y=max(criteria_long$value), 
           label="AIC", hjust=-0.1, color="purple") +
  labs(title = paste("Model Selection (Alpha =", best_alpha, ")"), y = "Standardized Score") +
  theme_minimal()

print(p1)
```

Gene count at each optimal log Lambda (AIC, BIC or EBIC)

```{r}
get_gene_count <- function(fit, lambda_val) {
  coefs <- coef(fit, s = lambda_val)
  sum(as.matrix(coefs) != 0) - 1 # Minus intercept
}

cat("Genes at AIC:", get_gene_count(full_fit, opt_lambda_aic), "\n")
cat("Genes at BIC:", get_gene_count(full_fit, opt_lambda_bic), "\n")
cat("Genes at EBIC:", get_gene_count(full_fit, opt_lambda_ebic), "\n")
```

Going with BIC, but increasing the BIC threshold to 24 (25 genes including target) using stability selection at the optimal BIC lambda and 1000, n/2 bootstraps for FDR control as described by Meinshausen & Bühlmann, 2010. This is because other approaches are either too sparse (EBIC) or large (AIC)

```{r}
n_stab_boot <- 10000
n_cores <- detectCores() - 1
cl <- makeCluster(n_cores)
registerDoParallel(cl)

# Subsampling function
run_lasso_sub <- function(x, y, lambdas, alpha_val) {
  sub_idx <- sample(1:nrow(x), floor(nrow(x))/2, replace = FALSE)
  fit <- glmnet(x[sub_idx,], y[sub_idx], alpha = alpha_val, lambda = lambdas)
  coefs <- coef(fit, s = lambdas)
  return(as.matrix(coefs[-1, ] != 0) * 1) 
}

# Export Data + functions to cluster
clusterExport(cl, c("x", "y", "lambda_seq", "run_lasso_sub", "best_alpha"))
clusterEvalQ(cl, library(glmnet))

# Bootstrap
message(paste("Running", n_stab_boot, "stability bootstraps..."))
stab_list <- foreach(i = 1:n_stab_boot) %dopar% {
  run_lasso_sub(x, y, lambda_seq, best_alpha)
}
stopCluster(cl)

# Stability probabilities
prob_matrix <- Reduce("+", stab_list) / n_stab_boot
colnames(prob_matrix) <- paste0("L", 1:length(lambda_seq))

# Column index at optimal BIC Lambda
target_lambda_idx <- which.min(abs(lambda_seq - opt_lambda_bic))

# Stability scores at that Lambda
gene_stability_scores <- prob_matrix[, target_lambda_idx]

# Create and sort df
stability_df <- data.frame(
  Gene = names(gene_stability_scores),
  Stability_Prob = as.numeric(gene_stability_scores)
)

# Sort descending
stability_df <- stability_df[order(-stability_df$Stability_Prob), ]

# Top 25 genes
top_genes <- head(stability_df$Gene, 25)

# Include HLA-DQB1
final_genes <- unique(c("HLA-DQB1", top_genes))

# Remove the weakest gene after adding HLA-DQB1 back in
if(length(final_genes) > 25) {
  final_genes <- final_genes[1:25]
}

# Save 
X_asthma_stable <- X_asthma[, final_genes]
write.csv(X_asthma_stable, "data/X_asthma_stable_25.csv", row.names = FALSE)

X_normal_stable <- X_normal[, final_genes]
write.csv(X_normal_stable, "data/X_normal_stable_25.csv", row.names = FALSE)
```

Plot PAGs 

```{r}
analyze_pag_sensitivity <- function(data, title = NULL, file_name = NULL) {
  
  # Filename setup
  data_arg_name <- deparse(substitute(data)) 
  
  if (is.null(file_name)) {
    clean_title <- gsub("[^[:alnum:]]", "_", title)
    file_name <- paste0(clean_title, ".pdf")
  }
  
  # Define alphas
  alphas <- c(0.01, 0.05, 0.10)
  
  # Initialize results df
  results_df <- data.frame(Alpha = numeric(), 
                            Edge_Count = integer(), 
                            Time_Sec = numeric())
  
  # suffStat
  suffStat <- list(C = cor(data, use = "pairwise.complete.obs"), n = nrow(data))
  
  pdf(file_name, width = 10, height = 10)
  
  for (alpha in alphas) {
    
    message(paste("Running FCI for Alpha:", alpha, "..."))
    
    # system.time
    pag.est <- NULL
    
    timer <- system.time({
      pag.est <- fci(suffStat, indepTest = gaussCItest, alpha = alpha, 
                     labels = colnames(data), verbose = FALSE)
    })
    
    # Extract elapsed time
    duration <- timer["elapsed"]
    
    # Count Edges
    amat <- pag.est@amat
    current_edges <- sum(amat[upper.tri(amat)] != 0)
    
    # Store Results
    results_df <- rbind(results_df, data.frame(Alpha = alpha, 
                                                 Edge_Count = current_edges, 
                                                 Time_Sec = as.numeric(duration)))
    
    # Plot PAG
    plot(pag.est)
    
    # Title
    title_text <- paste0(title, 
                         "\nalpha: ", alpha, 
                         "   edges: ", current_edges, 
                         "   time: ", round(duration, 3), "s")
    
    title(main = title_text, col.main = "blue", cex.main = 1.5)
  }
  
  dev.off()
  
  return(results_df)
}

results_asthma <- analyze_pag_sensitivity(X_asthma_top,
                                          file_name = "Asthma_PAG_Analysis.pdf")

results_normal <- analyze_pag_sensitivity(X_normal_top,
                                          file_name = "Normal_PAG_Analysis.pdf")

print(results_asthma)
print(results_normal)

# Visualize runtime
plot(results_asthma$Alpha, results_asthma$Time_Sec, 
     type = "b", pch = 19, col = "red",
     main = "Computation Time by Alpha",
     xlab = "Alpha", ylab = "Time (seconds)")

plot(results_normal$Alpha, results_normal$Time_Sec, 
     type = "b", pch = 19, col = "red",
     main = "Computation Time by Alpha",
     xlab = "Alpha", ylab = "Time (seconds)")
```

Use LV-IDA to see the causal effect of HLA-DQB1 on other genes

```{r}
library(pcalg)

# Ensure your function is loaded
source("lv-ida/lvida.R") 
source("lv-ida/iscyclic.R")

estimate_downstream_effects <- function(data, cause_var, n_boots = 500, title = NULL, file_name = NULL) {
  
  # Setup
  alphas <- c(0.01, 0.05, 0.10)
  cause_idx <- which(colnames(data) == cause_var)
  
  if (length(cause_idx) == 0) stop("Cause variable not found in data.")
  
  # "Predictors" here are actually potential TARGETS (Effects)
  potential_targets <- colnames(data)[-cause_idx]
  n_targets <- length(potential_targets)
  n_obs <- nrow(data)
  
  final_summary <- data.frame()
  
  # Title/File setup
  if (is.null(title)) title <- paste("Downstream Analysis | Cause:", cause_var)
  if (is.null(file_name)) {
    clean_cause <- gsub("[^[:alnum:]]", "_", cause_var)
    file_name <- paste0("Downstream_Effects_", clean_cause, ".pdf")
    csv_name <- paste0("Downstream_Effects_", clean_cause, ".csv")
  } else {
    # derive csv name from pdf name
    csv_name <- sub(".pdf$", ".csv", file_name)
  }
  
  pdf(file_name, width = 12, height = 10)
  
  for (alpha in alphas) {
    message(paste("\n=== Processing Alpha =", alpha, "==="))
    
    # Rows = Targets, Cols = Bootstrap Iterations
    boot_min_effects <- matrix(NA, nrow = n_targets, ncol = n_boots)
    boot_max_effects <- matrix(NA, nrow = n_targets, ncol = n_boots)
    
    # --- BOOTSTRAP LOOP ---
    for (b in 1:n_boots) {
      if (b %% 50 == 0) message(paste("  Bootstrap", b, "/", n_boots))
      
      # Resample
      boot_indices <- sample(1:n_obs, n_obs, replace = TRUE)
      boot_data <- data[boot_indices, ]
      
      # Re-calculate stats
      suffStat <- list(C = cor(boot_data, use = "complete.obs"), n = n_obs)
      mcov <- cov(boot_data, use = "complete.obs")
      
      # Run FCI
      pag.est <- tryCatch({
        fci(suffStat, indepTest = gaussCItest, alpha = alpha, labels = colnames(data), verbose = FALSE)
      }, error = function(e) return(NULL))
      
      if (is.null(pag.est)) next
      
      amat <- pag.est@amat
      
      # Loop through potential DOWNSTREAM targets
      for (i in seq_along(potential_targets)) {
        target_name <- potential_targets[i]
        target_idx <- which(colnames(data) == target_name)
        
        # --- LOGIC FLIP: Check if Cause -> Target is possible ---
        # We check if 'target_idx' is in the descendants of 'cause_idx'
        is_possible_descendant <- target_idx %in% pcalg::possibleDe(amat, cause_idx)
        
        if (!is_possible_descendant) {
          boot_min_effects[i, b] <- 0
          boot_max_effects[i, b] <- 0
        } else {
          
          # --- DYNAMIC FUNCTION CALL ---
          tryCatch({
            if (exists("lvida", mode = "function")) {
               # args: x (cause), y (effect), cov, graph
               effs <- lvida(cause_idx, target_idx, mcov, amat)
            } else {
               effs <- lvIda(cause_idx, target_idx, mcov, amat)
            }
            
            if (!all(is.na(effs))) {
              boot_min_effects[i, b] <- min(abs(effs))
              boot_max_effects[i, b] <- max(abs(effs))
            } else {
              boot_min_effects[i, b] <- 0
              boot_max_effects[i, b] <- 0
            }
          }, error = function(e) {
             boot_min_effects[i, b] <- NA
             boot_max_effects[i, b] <- NA
          })
        }
      }
    } 
    
    # Summarize Results for this alpha
    alpha_summary <- data.frame(
      Target_Gene = potential_targets,
      Alpha = alpha,
      Stability_Pct = rowMeans(boot_max_effects > 0, na.rm = TRUE) * 100,
      Median_Min_Effect = apply(boot_min_effects, 1, function(x) median(x[x > 0], na.rm = TRUE)),
      Median_Max_Effect = apply(boot_max_effects, 1, function(x) median(x[x > 0], na.rm = TRUE))
    )
    
    alpha_summary[is.na(alpha_summary)] <- 0
    final_summary <- rbind(final_summary, alpha_summary)
    
    # Reference Plot
    suffStat_full <- list(C = cor(data), n = nrow(data))
    pag_full <- fci(suffStat_full, indepTest = gaussCItest, alpha = alpha, labels = colnames(data))
    
    # Label top downstream targets
    top_targets <- alpha_summary[order(-alpha_summary$Stability_Pct), ][1:3, ]
    caption <- paste("Top Downstream Targets:", paste(top_targets$Target_Gene, "(", round(top_targets$Stability_Pct,0), "%)", collapse=", "))
    
    plot(pag_full, main = paste0(title, "\nalpha=", alpha, "\n", caption))
    
  } 
  
  dev.off()
  
  # --- NEW: Save to CSV ---
  write.csv(final_summary, csv_name, row.names = FALSE)
  message(paste("Saved edge list to:", csv_name))
  
  return(final_summary)
}

# --- Run ---
# Note: 'target_var' here is actually the CAUSE variable
results_asthma <- estimate_downstream_effects(X_asthma_HLAC, cause_var = "215175_at", n_boots = 500)
results_normal <- estimate_downstream_effects(X_normal_HLAC, cause_var = "215175_at", n_boots = 500)


```

